---
title: "Initial Posit Assistant Test — CHM Structural Metrics"
date: "2026-05-08"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

## Overview

This document records an initial test using Posit Assistant to compute
structural diversity metrics from a 25 cm canopy height model (CHM)
for the G2 site (`G2_25cm_chm.tif`). It details which metrics were
successfully computed, which required workarounds, and which failed.

---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(lidR)
library(terra)
library(tidyverse)
```

---

## Load & Inspect CHM

```{r load-chm}
chm_raw <- rast("../data/G2_25cm_chm.tif")
chm_raw
summary(values(chm_raw, na.rm = TRUE))
```

**Key findings from inspection:**

- Resolution: 0.25 m × 0.25 m (25 cm)
- Total cells: ~26.8 million (4470 × 5998)
- ~56% of cells are `NaN` — the raster extent covers a large non-study area
  (ocean, outside plot boundary, etc.)
- Only ~11.75 million cells contain valid data
- Minimum value is −11.88 m: negative values are noise artifacts from the
  point cloud processing pipeline and need to be clamped to 0

```{r clean-chm}
# Clamp negative noise artifacts to 0
chm <- clamp(chm_raw, lower = 0, values = TRUE)

pixel_res  <- res(chm)[1]   # 0.25 m
pixel_area <- pixel_res^2   # 0.0625 m²
ht_thresh  <- 2             # canopy / gap height threshold (m)

# Separate all valid pixels from canopy-only pixels
vals        <- values(chm, na.rm = TRUE)
canopy_vals <- vals[vals >= ht_thresh]

cat("Valid pixels:", length(vals), "\n")
cat("Canopy pixels (>= 2 m):", length(canopy_vals), "\n")
```

---

## ✅ Metrics That Worked

### Mean & Max Canopy Height

Calculated from canopy pixels only (≥ 2 m), after clamping negatives to 0.

```{r mean-max}
mean_ht <- mean(canopy_vals)
max_ht  <- max(canopy_vals)

cat("Mean canopy height:", round(mean_ht, 2), "m\n")
cat("Max canopy height: ", round(max_ht,  2), "m\n")
```

### Height Standard Deviation & Coefficient of Variation

```{r sd-cv}
sd_ht  <- sd(canopy_vals)
cv_ht  <- sd_ht / mean_ht

cat("Height SD:", round(sd_ht, 2), "m\n")
cat("CV:       ", round(cv_ht, 3), "\n")
```

### Height Percentiles

```{r percentiles}
pctls <- quantile(canopy_vals, probs = c(0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99))
round(pctls, 2)
```

### Canopy Cover & Gap Fraction

Computed over all valid pixels (includes gaps below 2 m threshold).

```{r cover-gap}
canopy_cover <- mean(vals >= ht_thresh)
gap_fraction <- 1 - canopy_cover

cat("Canopy cover (>= 2 m):", round(canopy_cover, 3),
    sprintf("(%.1f%%)\n", canopy_cover * 100))
cat("Gap fraction:         ", round(gap_fraction, 3),
    sprintf("(%.1f%%)\n", gap_fraction * 100))
```

### Canopy Rugosity

Defined here as the standard deviation of CHM surface heights across
canopy pixels — a simple measure of surface roughness.

```{r rugosity}
rugosity <- sd_ht
cat("Canopy rugosity (SD):", round(rugosity, 4), "m\n")
```

### Foliage Height Diversity (FHD)

Shannon entropy computed over 1 m height bins for canopy pixels. This is
an approximation — true FHD is typically derived from LAD profiles from
point cloud data, not a CHM. lidR's `entropy()` function was not used here
because it requires raw point cloud z-values rather than CHM surface heights.

```{r fhd}
breaks     <- seq(ht_thresh, ceiling(max_ht) + 1, by = 1)
bin_counts <- hist(canopy_vals, breaks = breaks, plot = FALSE)$counts
bin_props  <- bin_counts[bin_counts > 0] / sum(bin_counts)
fhd        <- -sum(bin_props * log(bin_props))

cat("Foliage Height Diversity:", round(fhd, 4), "nats\n")
```

### Vertical Complexity Index (VCI)

FHD normalised by `log(n_layers)` to scale between 0–1.

```{r vci}
vci <- fhd / log(length(bin_props))
cat("Vertical Complexity Index:", round(vci, 4), "\n")
```

### Canopy Gap Size Distribution

Gap pixels (< 2 m) identified as connected patches using `terra::patches()`.
This worked correctly but was **slow** (~2–3 minutes) due to the raster size
(~26.8M cells). Future runs should consider aggregating the raster or
clipping to the study area first.

```{r gap-dist, message=FALSE}
gap_mask    <- ifel(chm < ht_thresh, 1, NA)
gap_patches <- patches(gap_mask, directions = 8, zeroAsNA = TRUE)
gap_freq    <- freq(gap_patches)
gap_areas   <- sort(gap_freq$count * pixel_area)  # convert pixels → m²

tibble(
  n_gaps        = length(gap_areas),
  mean_gap_m2   = round(mean(gap_areas),   2),
  median_gap_m2 = round(median(gap_areas), 2),
  max_gap_m2    = round(max(gap_areas),    2),
  total_gap_m2  = round(sum(gap_areas),    2)
)
```

The gap size distribution is highly right-skewed: median gap is a single
pixel (0.06 m²), but there is one large gap of ~12,597 m² — likely the
main non-forested area within the study boundary.

---

## ❌ Metrics That Failed

### Rumple Index (`lidR::rumple_index()`)

**Status: Failed — returned `NaN`**

The `rumple_index()` function from lidR computes the ratio of 3D canopy
surface area to 2D ground area using Jenness's algorithm for raster inputs
(or Delaunay triangulation for point inputs).

Three approaches were attempted:

1. **SpatRaster input** — `rumple_index(chm)` → `NaN`
2. **Matrix input** — `rumple_index(as.matrix(chm, wide = TRUE), res = pixel_res)` → blank/`NaN`
3. **x, y, z vectors from a cropped area** — the crop accidentally landed
   entirely within the NaN region (56% of the raster is NaN), returning 0
   valid points

The likely root cause is the large proportion of NaN cells. Extracting all
valid x, y, z coordinates to compute via point triangulation was attempted
but **crashed the R session** — converting 11.75M points to a data frame
with `as.data.frame(chm, xy = TRUE, na.rm = TRUE)` was too memory-intensive.

**Recommended fix for next session:**

1. Clip the CHM to a valid bounding box first (remove the large NaN border)
2. Either pass the clipped raster to `rumple_index()`, or sample a
   manageable number of points:

```r
# Clip to valid extent, then compute
valid_ext  <- ext(trim(chm))
chm_clipped <- crop(chm, valid_ext)
ri <- rumple_index(chm_clipped)
```

---

## Summary Table

```{r summary-table, echo=FALSE}
tibble(
  Metric                       = c(
    "Mean canopy height",
    "Max canopy height",
    "Height SD",
    "Coefficient of variation",
    "P10 / P25 / P50 / P75 / P90",
    "Canopy cover",
    "Gap fraction",
    "Canopy rugosity",
    "Foliage Height Diversity",
    "Vertical Complexity Index",
    "Gap size distribution",
    "Rumple index"
  ),
  Status                       = c(
    "✅", "✅", "✅", "✅", "✅",
    "✅", "✅", "✅", "✅", "✅",
    "✅ (slow)", "❌ NaN"
  ),
  Value                        = c(
    paste(round(mean_ht, 2), "m"),
    paste(round(max_ht,  2), "m"),
    paste(round(sd_ht,   2), "m"),
    round(cv_ht, 3),
    paste(round(pctls[c("10%","25%","50%","75%","90%")], 1), collapse = " / "),
    sprintf("%.1f%%", canopy_cover * 100),
    sprintf("%.1f%%", gap_fraction * 100),
    paste(round(rugosity, 2), "m"),
    round(fhd, 4),
    round(vci, 4),
    "See gap table above",
    "NaN — see notes"
  )
) |>
  knitr::kable()
```

---

## Notes for Next Steps

- **Clip CHM to valid extent** before any heavy raster operations —
  the ~56% NaN border is causing failures and unnecessary computation.
- **Rumple index** should work once the raster is clipped; test with
  `terra::trim()` to remove the NaN border.
- **True FHD / LAD profiles** require the raw point cloud (`.las`/`.laz`),
  not just the CHM — the CHM-based FHD here is an approximation.
- Consider **aggregating to 1 m resolution** before `patches()` to speed
  up gap size distribution computation.
